---
title: "Google Data Analyst Capstone"
output: html_document
author: Avi Ghosh
---
## Main business question: How to convert rec bike users into members?
The goal initially is to point out how differently casual and members use the bike services. Using this information, advertisement and marketing campaigns can be put in place that would strategize to convert the appropriate segment of people. 

## Include the correct packages & libraries. 
First we load the required libraries

```{r install required packages, load libraries}
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(ggthemes)
```

## Set up your directory and load the dataset

```{r import the data: imported through dashboard}
setwd("~/Desktop/Google Data Analysis/casestudy")
feb21 <- read.csv("feb21.csv")
colnames(feb21)
```

## Conduct some initial analysis
Some of the analyses code have been commented out for brevity. 
Here I am identifying how the data is organized, sorting and filtering the data as needed, cleaning the data and getting used to the information provided. 

```{r prepare for analysis & clean the data, warning=FALSE}
dim(feb21)
#head(feb21)
#tail(feb21)
#str(feb21)
#summary(feb21)
#na.omit(feb21) # goes from 49k to 42k. have not
# assigned to the dataframe yet. 

unique(feb21$member_casual) #check for unique values
# only 2 values in the column, which is good. 
table(feb21$member_casual)#checknumber of obs.
# additional column: day of the week and ride duration
# have been added using Google Sheets. 

# add another column with ride duration in seconds
feb21$seconds <- difftime(feb21$ended_at, feb21$started_at)
str(feb21$seconds) #it is already numeric, no need to convert

nrow(feb21[feb21$seconds<0,]) #checking for negative values
nrow(feb21[feb21$seconds == 0,]) # 4 values with 0 duration
nrow(feb21[feb21$start_station_name == "HQ QR",])
#above bikes were taken out of docks and checked for quality
#they should be taken out. 
# we subset our dataset where station name was HQ QR or if trip duration was 0. 
feb21mod <- feb21[!(feb21$start_station_name == "HQ QR" | feb21$seconds == 0),]
sum(is.na(feb21mod$seconds))
```

We check the content of our dataset and subset/replace wherever deemed necessary. Next we will do some additional tests to check if our data
is skewed. There is a good chance it is. 

```{r Descriptive Stats 1}
mean(feb21mod$seconds, na.rm = TRUE)
median(feb21mod$seconds, na.rm = TRUE)
max(feb21mod$seconds, na.rm = TRUE)
min(feb21mod$seconds, na.rm = TRUE)
```

The mean is very far from median, this is often the case with 
mean, which is skewed by extreme values. I am going to fix this. 

```{r Descriptive Stats 2, warning=FALSE}
# take note of the outliers. 
quan_tile <- quantile(feb21mod$seconds, probs=c(.25, .75), na.rm = TRUE)
iqr <- IQR(feb21mod$seconds, na.rm=TRUE)
uprange <- quan_tile[2]+1.5*iqr
lowrange <- quan_tile[1]-1.5*iqr
print(quan_tile)
print(uprange)
print(lowrange)
feb21inrange <- subset(feb21mod, feb21mod$seconds > (quan_tile[1] - 1.5*iqr) 
          & feb21mod$seconds < (quan_tile[2]+1.5*iqr))
```

The new dataset only contains data excluding extreme values
skewing the dataset. Let us check out this altered dataset. 

```{r Descriptive Stats 3}
mean(feb21inrange$seconds, na.rm = TRUE)
median(feb21inrange$seconds, na.rm = TRUE)
max(feb21inrange$seconds, na.rm = TRUE)
min(feb21inrange$seconds, na.rm = TRUE)
```

Now the mean is much closer to the median. Using the IQR judiciously, extreme data points can be ignored. 

```{r Descriptive Stats 4, warning=FALSE}
# start comparing the different users. 
aggregate(feb21inrange$seconds ~ feb21inrange$member_casual, FUN = mean, na.rm = TRUE)
aggregate(feb21inrange$seconds ~ feb21inrange$member_casual, FUN = median, na.rm = TRUE)
aggregate(feb21inrange$seconds ~ feb21inrange$member_casual, FUN = max, na.rm = TRUE)
aggregate(feb21inrange$seconds ~ feb21inrange$member_casual, FUN = min, na.rm = TRUE)

# name the days of week replacing numbers to actual day names.
feb21inrange$day_of_week <- recode(feb21inrange$day_of_week, "1"="Sunday", "2"="Monday",
                        "3"="Tuesday",
                        "4"="Wednesday",
                        "5"="Thursday",
                        "6"="Friday",
                        "7"="Saturday")

# additionally breaking it down by day
aggregate(feb21inrange$seconds ~ feb21inrange$member_casual + feb21inrange$day_of_week, FUN = mean)

# analyze by type and weekday
feb21inrange %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()       #calculates the number of rides and average duration 
            ,average_duration = mean(seconds)) %>%   # calculates the average duration
  arrange(member_casual, weekday)
```

## Graphs

Now that we have aggregated our data, organized it, performed some calculations, we can visualize our dataset to get a more comprehensive picture

```{r visualize, warning=FALSE}
# visualization of number of rides
feb21inrange %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(seconds())) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + theme_wsj() + ggtitle("Number of Rides") +
  labs(fill = "User type") + theme(plot.title = element_text(hjust = 0.5))

# visualization for average duration
feb21inrange %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(seconds)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + theme_wsj() +
  theme_wsj() + ggtitle("Average Ride Duration(sec)") +
  labs(fill = "User type") + theme(plot.title = element_text(hjust = 0.5))

```

Some Observations: 
We notice that total number of rides is dominated by members throughout the week. If we look at the numbers of casual riders though, there is a significant spike on Saturday. Marketers could utilize this fact and have more presence in target locations on Saturday 
Secondly we also notice that casual members ride the bike longer than members consistently over all 7 days of the week. Further analysis can be done whether this difference is statistically significant. If found to be so, cost adjustments could be made so that a membership is more attractive for longer (casual) bike riders. 
